---
title: "InMODELA HIV-NCD Emulator development"
author: "Sharon Sawe"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup}

#Clean the working environment: Ensure you have saved all other things in the working environment that you may need in future before running this command
rm(list = ls())

knitr::opts_chunk$set(echo = T,message = FALSE,warning = FALSE, error = FALSE)
#install.packages("pacman",dependencies = T)
pacman::p_load(ggplot2,GauPro,bayesplot,rstanarm,dplyr,knitr)
#needed libraries
# library(ggplot2)
# library(Gaupro)
# library(bayesplot)
# library(rstanarm)

```

# Emulator development
We developed an emulator for HIV, hypertension and depression using the gaussian process

# Data
We used data from established simulators for HIV and the aforementioned NCDs which were built using data from Kenya.

## Gaussian Process
A Gaussian process fits a model to a dataset, which gives a function that gives a prediction for the mean at any point along with a variance of this prediction.
We implemented the gaussian process using the GauPro package.

## GauPro package
This R package provides R code for fitting Gaussian process models to data.

A detailed guide of the GauPro package in R by Collin Erickson can be found here:

https://cran.r-project.org/web/packages/GauPro/vignettes/GauPro.html

1. HIV

```{r}
# HIV modelling
# Starting with prevalence of HIV in Western Kenya
# The prevalence in 2018 is 5%
# The prevalence in 2028 is 3%

# x is the years
x <- seq(2018,2028,l=11)

# y is the prevalences for the two points in 2018 and 2028 whereby the seq will simulate the prevalence values in between 2018 and 2028
y <- seq(0.0481,0.0255, l=11)

plot(x, y,
     ylim=c(0,0.2), xlab="years", ylab= "HIV Prevalence")
# Linear model
lm_mod <- lm(y ~ x)
plot(x, y, xlab="years", ylab= "HIV Prevalence",type = "l",ylim = c(0,0.05))
abline(a=lm_mod$coef[1], b=lm_mod$coef[2], col='red')


# GAUPRO MODELLING
# A very useful property of GP's is that they give a predicted error. 
# The blue lines give an approximate 95% confidence interval.
# The width of the prediction interval is largest between points 
# and goes to zero near data points, which is what we would hope for.
l = c(0.0419,0.0395,0.0373,0.0351,0.0330,0.0309,0.0290,0.0271,0.0254,0.0238,0.0220)
u = c(0.0543,0.0512,0.0483,0.0455,0.0430,0.0402,0.0378,0.0354,0.0333,0.0312,0.0290
)
gp <- GauPro(x,y, parallel=FALSE)
plot(x,y, xlab="years", ylab= "HIV Prevalence", main= "Projected prevalence of HIV",type="b", lwd=2)#,
curve(gp$predict(x), add=T, col=2)
curve(gp$predict(x)+1.96*gp$predict(x, se=T)$se, add=T, col=4)
curve(gp$predict(x)-1.96*gp$predict(x, se=T)$se, add=T, col=4)
lines(x,l,col="yellow",lwd=4)
lines(x,u,col="yellow",lwd=4)

# Creating confidence intervals
# Create a dataframe using the annual emulator prevalence predictions
set.seed(1234)
new_x <- seq(min(x), max(x), length.out = 11) # 11 observations
predict_df11       <- predict(gp, new_x, se.fit = TRUE)
predict_df11$x     <- new_x
predict_df11$y     <- predict_df11$mean
predict_df11$lower <- predict_df11$y - 1.96 * predict_df11$se
predict_df11$upper <- predict_df11$y + 1.96 * predict_df11$se
View(predict_df11)

# 21 observations to obtain confidence intervals between years 
set.seed(1234)
new_x <- seq(min(x), max(x), length.out = 21) # 21 observations
predict_df21       <- predict(gp, new_x, se.fit = TRUE)
predict_df21$x     <- new_x
predict_df21$y     <- predict_df21$mean
predict_df21$lower <- predict_df21$y - 1.96 * predict_df21$se
predict_df21$upper <- predict_df21$y + 1.96 * predict_df21$se
View(predict_df21)
```

2. Hypertension

```{r,warning=FALSE,message=FALSE}
# Hypertension modelling
# The prevalence in 2018 is 29%
# The prevalence in 2028 is 35%

# x is the years
x <- seq(2018,2028,l=11)

# y is the prevalences for the two points in 2018 and 2028 whereby the seq will simulate the prevalence values in between 2018 and 2028
y <- seq(0.2947,0.3430, l=11)

# Plotting the linear regression
plot(x, y,
     ylim=c(0,0.4), xlab="years", ylab= "Hypertension Prevalence")
# Linear model
lm_mod <- lm(y ~ x)
plot(x, y, xlab="years", ylab= "Hypertension Prevalence",type = "l",ylim = c(0,0.4))
abline(a=lm_mod$coef[1], b=lm_mod$coef[2], col='red')


# GAUPRO MODELLING
# A very useful property of GP's is that they give a predicted error. 
# The blue lines give an approximate 95% confidence interval.
# The width of the prediction interval is largest between points 
# and goes to zero near data points, which is what we would hope for.
l = c(0.2845,0.2908,0.2965,0.3017,0.3073,0.3125,0.3171,0.3213,0.3250,0.3284,0.3313)
u = c(0.3050,0.3116,0.3177,0.3232,0.3292,0.3347,0.3396,0.3441,0.3480,0.3516,0.3547)
# Plotting
gp <- GauPro(x,y, parallel=FALSE)
plot(x,y, xlab="years", ylab= "Hypertension Prevalence", main= "Projected prevalence of Hypertension",type="b", lwd=2)#,
curve(gp$predict(x), add=T, col=2)
curve(gp$predict(x)+1.96*gp$predict(x, se=T)$se, add=T, col=4)
curve(gp$predict(x)-1.96*gp$predict(x, se=T)$se, add=T, col=4)
lines(x,l,col="yellow",lwd=4)
lines(x,u,col="yellow",lwd=4)

```

3. Comorbid HIV and Hypertension

```{r}
# Comorbid HIV and Hypertension modelling
# The prevalence in 2018 is 2.06%
# The prevalence in 2028 is 1.31%

# x is the years
x <- seq(2018,2028,l=11)

# y is the prevalences for the two points in 2018 and 2028 whereby the seq will simulate the prevalence values in between 2018 and 2028
y <- seq(0.0206,0.0131, l=11)
# Plotting the linear regression
plot(x, y,
     ylim=c(0,0.2), xlab="years", ylab= "HIV & Hypertension Prevalence")
# Linear model
lm_mod <- lm(y ~ x)
plot(x, y, xlab="years", ylab= "HIV & Hypertension Prevalence",type = "l",ylim = c(0,0.03))
abline(a=lm_mod$coef[1], b=lm_mod$coef[2], col='red')


# GAUPRO MODELLING
# A very useful property of GP's is that they give a predicted error. 
# The blue lines give an approximate 95% confidence interval.
# The width of the prediction interval is largest between points 
# and goes to zero near data points, which is what we would hope for.
l = c(0.0179,0.0173,0.0164,0.0158,0.0151,0.0144,0.0138,0.0132,0.0126,0.0120,0.0114)
u = c(0.0232,0.0224,0.0212,0.0203,0.0195,0.0186,0.0179,0.0170,0.0163,0.0156,0.0148)
# Plotting
gp <- GauPro(x,y, parallel=FALSE)
gp
plot(x,y, xlab="years", ylab= "HIV & Hypertension Prevalence", main= "Projected prevalence of HIV & Hypertension",type="b", lwd=2)#,
curve(gp$predict(x), add=T, col=2)
curve(gp$predict(x)+1.96*gp$predict(x, se=T)$se, add=T, col=4)
curve(gp$predict(x)-1.96*gp$predict(x, se=T)$se, add=T, col=4)
lines(x,l,col="yellow",lwd=4)
lines(x,u,col="yellow",lwd=4)
```

4. Depression

```{r}
# Depression modelling
# The prevalence in 2018 is 3.9%
# The prevalence in 2028 is 3.7%

# x is the years
x <- seq(2018,2028,l=11)

# y is the prevalences for the two points in 2018 and 2028 whereby the seq will simulate the prevalence values in between 2018 and 2028
y <- seq(0.039,0.037, l=11)

# Plotting the linear regression
plot(x, y,
     ylim=c(0,0.04), xlab="years", ylab= "Depression Prevalence")
# Linear model
lm_mod <- lm(y ~ x)
plot(x, y, xlab="years", ylab= "Depression Prevalence",type = "l",ylim = c(0,0.04))
abline(a=lm_mod$coef[1], b=lm_mod$coef[2], col='red')


# GAUPRO MODELLING
# A very useful property of GP's is that they give a predicted error. 
# The blue lines give an approximate 95% confidence interval.
# The width of the prediction interval is largest between points 
# and goes to zero near data points, which is what we would hope for.

# Plotting
gp <- GauPro(x,y, parallel=FALSE)
gp
plot(x,y, xlab="years", ylab= "Depression Prevalence", main= "Projected prevalence of Depression",type="b", lwd=2)#,
curve(gp$predict(x), add=T, col=2)
curve(gp$predict(x)+1.96*gp$predict(x, se=T)$se, add=T, col=4)
curve(gp$predict(x)-1.96*gp$predict(x, se=T)$se, add=T, col=4)

```

```{r}
# COMBINING THE PLOTS


line = 1
cex = 1
side = 3
adj=-0.05

par(mfrow=c(2,2), oma=c(1,6,1,1))


### PLOT 1
# HIV
x <- seq(2018,2028, l=11)
y <- seq(0.0481,0.0255, l=11)
# The simulator's confidence intervals
l = c(0.0419,0.0395,0.0373,0.0351,0.0330,0.0309,0.0290,0.0271,0.0254,0.0238,0.0220)
u = c(0.0543,0.0512,0.0483,0.0455,0.0430,0.0402,0.0378,0.0354,0.0333,0.0312,0.0290
)
# Fitting the gaussian process
library(GauPro)
gp <- GauPro(x,y, parallel=FALSE)
# Plotting
plot(x,y,xlab="years", ylab= "HIV Prevalence",type="b", lwd=2, pch=16)#,
curve(gp$predict(x), add=T, col=2, lty=5)
curve(gp$predict(x)+1.96*gp$predict(x, se=T)$se, add=T, col=4)
curve(gp$predict(x)-1.96*gp$predict(x, se=T)$se, add=T, col=4)
lines(x,l,col="yellow",lwd=4)
lines(x,u,col="yellow",lwd=4)
mtext("A", side=side, line=line, cex=cex, adj=adj)


# PLOT 2
# Hypertension
x <- seq(2018,2028, l=11)
y <- seq(0.2947,0.3430, l=11)
# Fitting the gaussian process
# The simulator's confidence intervals
l = c(0.2845,0.2908,0.2965,0.3017,0.3073,0.3125,0.3171,0.3213,0.3250,0.3284,0.3313)
u = c(0.3050,0.3116,0.3177,0.3232,0.3292,0.3347,0.3396,0.3441,0.3480,0.3516,0.3547)
# Fitting the gaussian process
library(GauPro)
gp <- GauPro(x,y, parallel=FALSE)
# Plotting
plot(x,y, xlab="years", ylab= "Hypertension Prevalence",type="b", lwd=2, pch=16)#,
curve(gp$predict(x), add=T, col=2, lty=5)
curve(gp$predict(x)+1.96*gp$predict(x, se=T)$se, add=T, col=4)
curve(gp$predict(x)-1.96*gp$predict(x, se=T)$se, add=T, col=4)
lines(x,l,col="yellow",lwd=4)
lines(x,u,col="yellow",lwd=4)
mtext("B", side=side, line=line, cex=cex, adj=adj)

# PLOT 3
# Comorbid HIV and hypertension
x <- seq(2018,2028, l=11)
y <- seq(0.0206,0.0131, l=11)
# Fitting the gaussian process
# The simulator's confidence intervals
l = c(0.0179,0.0173,0.0164,0.0158,0.0151,0.0144,0.0138,0.0132,0.0126,0.0120,0.0114)
u = c(0.0232,0.0224,0.0212,0.0203,0.0195,0.0186,0.0179,0.0170,0.0163,0.0156,0.0148)
# Fitting the gaussian process
library(GauPro)
gp <- GauPro(x,y, parallel=FALSE)
# Plotting
plot(x,y, xlab="years", ylab= "HIV & Hypertension Prevalence",type="b", lwd=2, pch=16)#,
curve(gp$predict(x), add=T, col=2, lty=5)
curve(gp$predict(x)+1.96*gp$predict(x, se=T)$se, add=T, col=4)
curve(gp$predict(x)-1.96*gp$predict(x, se=T)$se, add=T, col=4)
lines(x,l,col="yellow",lwd=4)
lines(x,u,col="yellow",lwd=4)
mtext("C", side=side, line=line, cex=cex, adj=adj)

# PLOT 4
x <- seq(2018,2028, l=11)
y <- seq(0.039,0.037, l=11)
# Fitting the gaussian process
library(GauPro)
gp <- GauPro(x,y, parallel=FALSE)
# Plotting
plot(x,y, xlab="years", ylab= "Depression Prevalence",type="b", lwd=2, pch=16)
curve(gp$predict(x), add=T, col = 2, lty=5)
curve(gp$predict(x)+1.96*gp$predict(x, se=T)$se, add=T, col = 4)
curve(gp$predict(x)-1.96*gp$predict(x, se=T)$se, add=T, col = 4)
mtext("D", side=side, line=line, cex=cex, adj=adj)
```
# BAYESIAN POSTERIOR PREDICTIVE ANALYSIS
- This step provides posterior predictive distributions that can be used to analyze the credibility of future observable data based on the posterior distribution.
- The procedure is implemented via the Bayesian inference rstanarm package
- A detailed guide of the posterior predictive analysis can be found here:

https://cran.r-project.org/web/packages/bayesplot/vignettes/graphical-ppcs.html
```{r}
# Combination
side=3
line = 1
cex = 1
adj=-0.05

par(mfrow=c(2,2), oma=c(1,6,1,1))

set.seed(12345)


# PLOT 1
# HIV
x <- c(2018,2019,2020,2021,2022,2023,2024,2025,2026,2027,2028)
y <- c(0.0481,0.0454,0.0428,0.0403,0.0380,0.0355,0.0334,0.0313,0.0294,0.0275,0.0255)
# R stanarm
sm1   <- stan_glm(y~x, family=gaussian(), data = data.frame(x,y), seed=1234)
# Y predicted
ypred <- posterior_predict(sm1, draws = 50)
# density plot
plot(density(y),col="black",lwd=3, xlab="HIV prevalence", main="HIV")
lines(density(ypred[1,]), col="red", lty=2, lwd=3)
for(i in 1:nrow(ypred))
{lines(density(ypred[i,]),col="pink",lty=3)}
#lines(density(colMeans(ypred)), col="red", lty=2, lwd=3)
lines(density(ypred[1,]), col="red", lty=2, lwd=3)
legend("bottomright", c("y", "post-pred y", "y rep"), lty=c(1,2,3), lwd=2, col=c("black", "red", "pink"), bty="n")
mtext("A", side=side, line=line, cex=cex, adj=adj)




# PLOT 2
# Hypertension
x <- c(2018,2019,2020,2021,2022,2023,2024,2025,2026,2027,2028)
y <- c(0.2947,0.3012,0.3071,0.3125,0.3182,0.3236,0.3283,0.3327,0.3365,0.3400,0.3430)
# R stanam
sm2   <- stan_glm(y~x, family=gaussian(), data = data.frame(x,y), seed=1234)
# Y predicted
ypred <- posterior_predict(sm2, draws = 50)
# density plot
plot(density(y),col="black", lwd=3, xlab="Hypertension prevalence", main="Hypertension")
lines(density(ypred[1,]), col="red", lty=2, lwd=3)
for(i in 1:nrow(ypred))
{lines(density(ypred[i,]),col="pink",lty=3)}
#lines(density(colMeans(ypred)), col="red", lty=2, lwd=3)
lines(density(ypred[1,]), col="red", lty=2, lwd=3)
legend("bottomright", c("y", "post-pred y", "y rep"), lty=c(1,2,3), lwd=2, col=c("black", "red", "pink"), bty="n")
mtext("B", side=side, line=line, cex=cex, adj=adj)



# PLOT 3
# Comorbid HIV and hypertension
x <- c(2018,2019,2020,2021,2022,2023,2024,2025,2026,2027,2028)
y <- c(0.0206,0.0198,0.0188,0.0181,0.0173,0.0165,0.0158,0.0151,0.0144,0.0138,0.0131)
# R stanam
sm3   <- stan_glm(y~x, family=gaussian(), data = data.frame(x,y), seed=1234)
# Y predicted
ypred <- posterior_predict(sm3, draws = 50)
# density plot
plot(density(y), col="black", lwd=3, xlab="HIV & Hypertension prevalence", main="HIV & Hypertension")
lines(density(ypred[1,]), col="red", lty=2, lwd=3)
for(i in 1:nrow(ypred))
{lines(density(ypred[i,]), col="pink", lty=3)}
#lines(density(colMeans(ypred)), col="red", lty=2, lwd=3)
lines(density(ypred[1,]), col="red", lty=2, lwd=3)
legend("bottomright", c("y", "post-pred y", "y rep"), lty=c(1,2,3), lwd=2, col=c("black", "red", "pink"), bty="n")
mtext("C", side=side, line=line, cex=cex, adj=adj)


# PLOT 4
# Depression
x <- c(2018,2020,2025,2030,2035)
y <- c(0.039,0.038,0.037,0.037,0.036)
# R stanam
sm4   <- stan_glm(y~x, family=gaussian(), data = data.frame(x,y), seed=1234)
# Y predicted
ypred <- posterior_predict(sm4, draws = 50)
# density plot
plot(density(y),col="black", lwd=3, xlab="Depression prevalence", main="Depression")
lines(density(ypred[1,]), col="red", lty=2, lwd=3)
for(i in 1:nrow(ypred))
{lines(density(ypred[i,]),col="pink",lty=3)}
#lines(density(colMeans(ypred)), col="red", lty=2, lwd=3)
lines(density(ypred[1,]), col="red", lty=2, lwd=3)
legend("bottomright", c("y", "post-pred y", "y rep"), lty=c(1,2,3), lwd=2, col=c("black", "red", "pink"), bty="n")
mtext("D", side=side, line=line, cex=cex, adj=adj)


```
# EMULATOR VALIDATION
- We evaluate and validate our emulator using the leave-one-out cross validation (LOO) technique
- The technique is implemented using the loo R package whereby the computation is done using Pareto-smoothed importance sampling (PSIS) which regularizes importance weights
- A detailed guide can be found here:
https://cran.r-project.org/web/packages/loo/vignettes/loo2-large-data.html

https://andrewproctor.github.io/rcourse/module6.html#model_testing

```{r}
# HIV model validation
# We call sm1 from the above posterior predictive analysis
# Getting the credible intervals
posterior_interval(sm1, prob=0.95)

# Graphical Credible intervals
plot(sm1)

# Juxtaposing the prior and the posterior
posterior_vs_prior(sm1)

# Graphical posterior predictive analysis
pp_check(sm1)
# Watanabe-Akaike information criterion
waic(sm1)
# Leave One Out Cross-Validation
loo(sm1)


# HYPERTENSION MODEL EVALUATION
# We call sm2 from the above posterior predictive analysis
# Getting the credible intervals
posterior_interval(sm2, prob=0.95)

# Graphical Credible intervals
plot(sm2)

# Juxtaposing the prior and the posterior
posterior_vs_prior(sm2)

# Graphical posterior predictive analysis
pp_check(sm2)
# Watanabe-Akaike information criterion
waic(sm2)
# Leave One Out Cross-Validation
loo(sm2)


# COMORBID HIV AND HYPERTENSION MODEL EVALUATION
# We call sm3 from the above posterior predictive analysis
# Getting the credible intervals
posterior_interval(sm3, prob=0.95)

# Graphical Credible intervals
plot(sm3)

# Juxtaposing the prior and the posterior
posterior_vs_prior(sm3)

# Graphical posterior predictive analysis
pp_check(sm3)
# Watanabe-Akaike information criterion
waic(sm3)
# Leave One Out Cross-Validation
loo(sm3, k_threshold = 0.7) #the threshold is put in order to calculate the ELPD without the assumption that these observations are negligible. 
# This will refit the model 1 times to compute the ELPDs for the problematic observations directly.


# DEPRESSION MODEL VALIDATION
# We call sm4 from the above posterior predictive analysis
# Getting the credible intervals
posterior_interval(sm4, prob=0.95)

# Graphical Credible intervals
plot(sm4)

# Juxtaposing the prior and the posterior
posterior_vs_prior(sm4)

# Graphical posterior predictive analysis
pp_check(sm4)
# Watanabe-Akaike information criterion
waic(sm4)
# Leave One Out Cross-Validation
loo(sm4)

```
# Loo output explanation   
- From the above loo cross validation outputs, we observe that all the four GP models have pareto k estimates with k-threshold values of k<0.7 which are okay.
-We therefore validate that our emulator is a good representation of the simulators.